% Mirror of docs/ocannl-formal-core-technical-report.md.
% Regenerate or update alongside the Markdown report when the proof text changes.
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{\textbf{#1}}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{The Formal Core of OCANNL Shape and Projection Inference}
\author{}
\date{}

\begin{document}

\maketitle

This report is a standalone companion to the OCANNL workshop article,
\texttt{docs/ocannl\_workshop\_article\_human.md}. It consolidates the
working notes in \texttt{docs/blog/ocannl-formal-core.md} and
\texttt{docs/formal-core-appendix.md} into one coherent account. The
scope is the core shape language used by the current proof effort:
dimensions, row variables, broadcasting, flat row equality, solving,
closing, and core projection inference. The staged extensions for affine
indexing, convolution padding, concatenation, and total-element
constraints are discussed only to delimit what is already part of the
core proof and what is outside that core.

The main result is deliberately not overstated. The core semantic rules
are proved sound; finite solving terminates with the persistent
rank-cycle guard; successful closing produces a real solution; and
projection inference over closed core shapes is canonical and
address-safe.

What is proved here:

\begin{itemize}
\tightlist
\item
  The rank-cycle guard is sound and exact for the facts it records.
\item
  The Detection Lemma is proved: every divergent rank-unsatisfiable run
  eventually records a positive cycle.
\item
  Projection inference is fully proved for the core projection language.
\end{itemize}

The remaining caveats are separate from those proved core results:

\begin{itemize}
\tightlist
\item
  The deterministic solver commits policy choices for marker placement
  and for deferred word equations. These choices are sound but
  incomplete for the abstract \texttt{approx} semantics. The conjecture
  that surface OCANNL programs do not observe the placement
  incompleteness remains open.
\item
  The affine, convolution, and concatenation strata used by the
  implementation need a separate staged soundness proof.
\end{itemize}

The implementation was checked against the formal rules while preparing
this report, especially \texttt{tensor/row.ml} and
\texttt{tensor/shape.ml}. The current code already implements the key
corrections discovered by the formalization: flat closed-row equality,
deferred shifted self equations/inequalities, explicit closed-closed
inequality checks across structural flanks, and the persistent rank
graph. No code change was required for this report.

\subsection{1. Dimensions}\label{1-dimensions}

Fix a set \texttt{B} of basis tags with a distinguished ordinary basis
\texttt{default}. There is also a special claim-free unit written
\texttt{1\_empty}; in the implementation it is represented by a
dimension of size \texttt{1} with basis \texttt{bcast\_if\_1}. The
mathematical presentation keeps it separate from ordinary atoms,
including the ordinary atom \texttt{1\_default}.

\textbf{Definition 1.1 (Dimensions).} The set of dimensions is

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{D = \{ 1\_empty \} union \{ n\_b | n \textgreater{}= 1 and b in B \}.}
\end{Highlighting}
\end{Shaded}

Elements \texttt{n\_b} are atoms. Adjoin a fresh bottom element
\texttt{bottom} and write
\texttt{D\_bottom\ =\ D\ union\ \{\ bottom\ \}}.

\textbf{Definition 1.2 (Broadcast order).} On \texttt{D},

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{d1 \textless{}= d2    iff    d2 = 1\_empty or d1 = d2.}
\end{Highlighting}
\end{Shaded}

Extend this to \texttt{D\_bottom} by \texttt{bottom\ \textless{}=\ d}
for every \texttt{d}.

Thus \texttt{1\_empty} is the greatest element, atoms are pairwise
incomparable, and \texttt{bottom} is only a proof device for meets.

\textbf{Proposition 1.3 (Dimension lattice).} \texttt{D\_bottom} is a
bounded lattice of height two. Its top is \texttt{1\_empty}, its bottom
is \texttt{bottom}, and the atoms form an antichain between them. Meets
and joins are:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{d meet d = d}
\NormalTok{d meet 1\_empty = d}
\NormalTok{d1 meet d2 = bottom        for distinct atoms d1, d2}

\NormalTok{d join d = d}
\NormalTok{d join bottom = d}
\NormalTok{d1 join d2 = 1\_empty       for distinct atoms d1, d2.}
\end{Highlighting}
\end{Shaded}

\textbf{Proof.} Reflexivity follows from equality in Definition 1.2. For
antisymmetry, if \texttt{d1\ \textless{}=\ d2} and
\texttt{d2\ \textless{}=\ d1} and \texttt{d1\ !=\ d2}, then the first
inequality forces \texttt{d2\ =\ 1\_empty}, while the second forces
\texttt{d1\ =\ 1\_empty}, a contradiction. For transitivity, if the
rightmost element is \texttt{1\_empty} the claim is immediate; otherwise
the second inequality is equality, and the first inequality gives the
result.

The displayed meet and join equations exhaust all cases: equal
arguments, a top or bottom argument, and two distinct atoms. In each
case the candidate is checked directly from the order. A distinct pair
of atoms has no common lower bound above \texttt{bottom} and no common
upper bound below \texttt{1\_empty}. Therefore the displayed operations
are the greatest lower bounds and least upper bounds. \texttt{square}

\textbf{Proposition 1.4 (Non-distributivity).} If there are at least
three atoms, \texttt{D\_bottom} is not distributive. It is, however,
modular; the earlier working claim that the lattice is not modular was
false.

\textbf{Proof.} Let \texttt{a}, \texttt{b}, and \texttt{c} be distinct
atoms. The sublattice \texttt{\{bottom,\ a,\ b,\ c,\ 1\_empty\}} is the
diamond lattice \texttt{M\_3}. It violates distributivity:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{a meet (b join c) = a meet 1\_empty = a}
\NormalTok{(a meet b) join (a meet c) = bottom join bottom = bottom.}
\end{Highlighting}
\end{Shaded}

So any lattice containing it as a sublattice is not distributive. The
same argument cannot show non-modularity because \texttt{M\_3} is
modular. In fact the height-two atom lattice is modular: if
\texttt{x\ \textless{}=\ z}, then the only non-trivial case is
\texttt{x\ =\ bottom} or \texttt{z\ =\ 1\_empty}, and the modular law
reduces to the defining equations for meet and join. \texttt{square}

\subsection{2. Rows}\label{2-rows}

A row is a sequence of axes with a distinguished insertion point. The
marker is not syntax trivia: it is where broadcasting inserts implicit
axes and where a row variable absorbs axes during inference.

\textbf{Definition 2.1 (Rows).} A ground row is a triple
\texttt{l\ .\ diamond\ .\ r} with \texttt{l,\ r\ in\ D*}. The lists are
the leading and trailing flanks. Its rank is
\texttt{\textbar{}l\textbar{}\ +\ \textbar{}r\textbar{}}. Row identity
is marker-sensitive:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{[3] . diamond . [4] != [3,4] . diamond . [].}
\end{Highlighting}
\end{Shaded}

\textbf{Definition 2.2 (Expansion).} For
\texttt{R\ =\ l\ .\ diamond\ .\ r} and
\texttt{n\ \textgreater{}=\ rank(R)},

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{R up n = l . 1\_empty\^{}(n {-} rank(R)) . r.}
\end{Highlighting}
\end{Shaded}

The expansion inserts claim-free broadcast axes at the marker.

\textbf{Definition 2.3 (Row order).} Write
\texttt{R2\ =\ l2\ .\ diamond\ .\ r2} and
\texttt{R1\ =\ l1\ .\ diamond\ .\ r1}. Then
\texttt{R2\ \textless{}=\ R1} iff:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \texttt{\textbar{}l1\textbar{}\ \textless{}=\ \textbar{}l2\textbar{}}
  and
  \texttt{\textbar{}r1\textbar{}\ \textless{}=\ \textbar{}r2\textbar{}};
\item
  for every position \texttt{i\ \textless{}\ rank(R2)},
  \texttt{(l2\ .\ r2){[}i{]}\ \textless{}=\ (R1\ up\ rank(R2)){[}i{]}}.
\end{enumerate}

The left row is the more material row. A shorter row can sit above a
longer one because the shorter row has more broadcast credit at its
marker.

\textbf{Lemma 2.4 (Expansion monotonicity).} If
\texttt{R2\ \textless{}=\ R1}, then for every
\texttt{n\ \textgreater{}=\ rank(R2)} and every position
\texttt{i\ \textless{}\ n},

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{(R2 up n)[i] \textless{}= (R1 up n)[i].}
\end{Highlighting}
\end{Shaded}

\textbf{Proof.} Write \texttt{Rj\ =\ lj\ .\ diamond\ .\ rj}. There are
three regions.

If \texttt{i\ \textless{}\ \textbar{}l2\textbar{}}, the left side is
\texttt{l2{[}i{]}}. When
\texttt{i\ \textless{}\ \textbar{}l1\textbar{}}, the defining pointwise
condition gives \texttt{l2{[}i{]}\ \textless{}=\ l1{[}i{]}}. When
\texttt{i\ \textgreater{}=\ \textbar{}l1\textbar{}}, the position lies
in the middle of \texttt{R1\ up\ n} unless it is already in the trailing
region; but
\texttt{i\ \textless{}\ \textbar{}l2\textbar{}\ \textless{}=\ n\ -\ \textbar{}r2\textbar{}\ \textless{}=\ n\ -\ \textbar{}r1\textbar{}},
so it is not trailing. Hence the right side is \texttt{1\_empty}, and
the inequality is automatic.

If
\texttt{\textbar{}l2\textbar{}\ \textless{}=\ i\ \textless{}\ n\ -\ \textbar{}r2\textbar{}},
the left side is \texttt{1\_empty}. By the flank-fit conditions,
\texttt{i} also lies in the middle of \texttt{R1\ up\ n}, so the right
side is \texttt{1\_empty}.

If \texttt{i\ \textgreater{}=\ n\ -\ \textbar{}r2\textbar{}}, the proof
is symmetric to the leading-flank case, using the outer-right alignment
of trailing flanks. \texttt{square}

\textbf{Proposition 2.5 (Partial order).} The row relation
\texttt{\textless{}=} is a partial order.

\textbf{Proof.} Reflexivity is immediate. For antisymmetry, mutual
flank-fit gives equal leading and trailing flank lengths. The expansions
at that common rank are the rows\textquotesingle{} flat contents, so
pointwise antisymmetry in \texttt{D} gives equal flanks.

For transitivity, suppose
\texttt{R3\ \textless{}=\ R2\ \textless{}=\ R1} and let
\texttt{n\ =\ rank(R3)}. Flank-fit conditions compose. The pointwise
condition follows from \texttt{R3\ up\ n\ \textless{}=\ R2\ up\ n} by
the first inequality and \texttt{R2\ up\ n\ \textless{}=\ R1\ up\ n} by
Lemma 2.4 applied to \texttt{R2\ \textless{}=\ R1}. Transitivity in
\texttt{D} completes the proof. \texttt{square}

\textbf{Proposition 2.6 (Top row).} The empty marked row
\texttt{{[}{]}\ .\ diamond\ .\ {[}{]}} is the greatest row, and it is
the unique greatest row.

\textbf{Proof.} Its flank lengths are zero and its expansion to any rank
is all \texttt{1\_empty}, so every row refines it. If some row is above
the empty row, the flank-fit inequalities force both of its flanks to
have length zero. \texttt{square}

\textbf{Proposition 2.7 (Row joins).} Rows form a join-semilattice with
top. The least upper bound of \texttt{R1\ =\ l1\ .\ diamond\ .\ r1} and
\texttt{R2\ =\ l2\ .\ diamond\ .\ r2} is
\texttt{Rj\ =\ lj\ .\ diamond\ .\ rj}, where:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{|lj| = min(|l1|, |l2|)}
\NormalTok{lj[i] = l1[i] join l2[i]}
\end{Highlighting}
\end{Shaded}

and similarly for the trailing flank, aligned from the right.

\textbf{Proof.} \texttt{Rj} is an upper bound because its flanks are no
longer than either input flank. On positions retained in \texttt{Rj},
the dimension join is an upper bound; positions outside the retained
flanks expand to \texttt{1\_empty}.

Now let \texttt{U} be any upper bound. Since
\texttt{R1\ \textless{}=\ U} and \texttt{R2\ \textless{}=\ U}, the
flanks of \texttt{U} are no longer than the corresponding flanks of both
inputs, hence no longer than the flanks of \texttt{Rj}. Thus the
flank-fit condition for \texttt{Rj\ \textless{}=\ U} holds. At each
retained leading position of \texttt{U}, both \texttt{l1{[}i{]}} and
\texttt{l2{[}i{]}} are below \texttt{U}\textquotesingle s dimension
there, so their join is below it. The trailing side is symmetric.
Positions in the middle of \texttt{U} are \texttt{1\_empty}. Therefore
\texttt{Rj\ \textless{}=\ U}. \texttt{square}

\textbf{Definition 2.8 (Flat equivalence).} The flat content of
\texttt{l\ .\ diamond\ .\ r} is \texttt{l\ .\ r}. Define
\texttt{R\ approx\ R\textquotesingle{}} iff their flat contents are
equal. Equivalently, they have equal rank and equal expansion at that
rank.

\texttt{approx} is not the identity of row elements. It is the equality
relation used for row equality constraints, including einsum template
matching.

\textbf{Proposition 2.9 (\texttt{approx} versus \texttt{\textless{}=}).}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The equivalence induced by mutual \texttt{\textless{}=} is identity.
\item
  Distinct rows related by \texttt{approx} are incomparable under
  \texttt{\textless{}=}.
\item
  \texttt{approx} is not a congruence for the marked row order.
\end{enumerate}

\textbf{Proof.} Item 1 is antisymmetry. For item 2, two flat-equal but
distinct rows must place the marker at different positions. If one has a
longer leading flank, it has a shorter trailing flank because total rank
is equal. The leading flank-fit condition fails in one direction and the
trailing flank-fit condition fails in the other.

For item 3, take

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{R  = [3] . diamond . [2,4]}
\NormalTok{S  = [3] . diamond . [4]}
\NormalTok{S\textquotesingle{} = [3,4] . diamond . [].}
\end{Highlighting}
\end{Shaded}

Then \texttt{S\ approx\ S\textquotesingle{}}, and
\texttt{R\ \textless{}=\ S}: the expansion of \texttt{S} to rank three
is \texttt{{[}3,\ 1\_empty,\ 4{]}}. But
\texttt{R\ \textless{}=\ S\textquotesingle{}} fails because the upper
row would need a leading flank of length at most one, while
\texttt{S\textquotesingle{}} has leading length two. Thus replacing an
\texttt{approx}-equivalent row inside an inequality can change truth.
\texttt{square}

This is the first non-principality point. Flat equality is needed for
useful spec matching, but marker placement remains a real policy choice
whenever an equality binds a row variable.

\subsubsection{2.1 Rigid Rows and One-Shot
Checking}\label{21-rigid-rows-and-one-shot-checking}

It is useful to name the marker-free interpretation of a flat row.

\textbf{Definition 2.10 (Rigid rows).} A rigid row is
\texttt{F\^{}bullet} for \texttt{F\ in\ D\^{}n}. Its expansion is
defined only at rank \texttt{n} and equals \texttt{F}. Extend the order
by:

\begin{itemize}
\tightlist
\item
  marked-marked: Definition 2.3;
\item
  rigid-rigid: equal rank and pointwise refinement;
\item
  rigid below marked: \texttt{F\^{}bullet\ \textless{}=\ R} iff
  \texttt{rank(F)\ \textgreater{}=\ rank(R)} and
  \texttt{F{[}i{]}\ \textless{}=\ (R\ up\ rank(F)){[}i{]}} pointwise;
\item
  marked below rigid: never.
\end{itemize}

\textbf{Proposition 2.11 (Two-sorted order).} The extended relation is a
partial order. The empty marked row remains the unique top. Rigid row
equality is flat content equality.

\textbf{Proof.} Reflexivity and antisymmetry hold within each sort as
before; a cross-sort mutual comparison is impossible because marked rows
are never below rigid rows. For transitivity, the all-marked case is
Proposition 2.5, the all-rigid case is pointwise transitivity, and the
mixed cases are:

\begin{itemize}
\tightlist
\item
  rigid \texttt{\textless{}=} rigid \texttt{\textless{}=} marked, where
  the equal rigid ranks and pointwise inequalities compose;
\item
  rigid \texttt{\textless{}=} marked \texttt{\textless{}=} marked, where
  Lemma 2.4 transports the second marked inequality to the rigid rank.
\end{itemize}

No other mixed chain is possible because marked-below-rigid comparisons
are forbidden. Topness of the empty marked row follows as in Proposition
2.6. \texttt{square}

The forbidden marked-below-rigid clause is forced. If even rank-equal
marked rows were allowed below their rigidifications, then

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{[5] . diamond . [3] \textless{}= [] . diamond . [3] \textless{}= [3]\^{}bullet}
\end{Highlighting}
\end{Shaded}

would require
\texttt{{[}5{]}\ .\ diamond\ .\ {[}3{]}\ \textless{}=\ {[}3{]}\^{}bullet},
impossible by rank.

\textbf{Definition 2.12 (One-shot order).} Write
\texttt{R\ \textless{}=1\ S} when
\texttt{flat(R)\^{}bullet\ \textless{}=\ S} in the two-sorted order.
This is the marker-erasing, single-check relation used to state results
that should not depend on the solver\textquotesingle s marker-placement
policy.

\subsection{3. Terms, Substitutions, and
Constraints}\label{3-terms-substitutions-and-constraints}

\textbf{Definition 3.1 (Terms).} Dimension terms are:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{t ::= alpha | n\_b | 1\_empty.}
\end{Highlighting}
\end{Shaded}

Row terms are:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{R ::= l . diamond . r | l . \textless{}rho\textgreater{} . r}
\end{Highlighting}
\end{Shaded}

where \texttt{l} and \texttt{r} are sequences of dimension terms. A term
is ground if it is variable-free. Each row term contains at most one row
variable, at the marker.

\textbf{Definition 3.2 (Substitution).} A substitution is a finite
sort-respecting map from variables to terms. It is kept idempotent:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dom(sigma) cap vars(range(sigma)) = empty.}
\end{Highlighting}
\end{Shaded}

Application to row variables splices at the marker. If
\texttt{sigma(rho)\ =\ l\textquotesingle{}\ .\ \textless{}rho\textquotesingle{}\textgreater{}\ .\ r\textquotesingle{}},
then

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sigma(l . \textless{}rho\textgreater{} . r) = sigma(l) . l\textquotesingle{} . \textless{}rho\textquotesingle{}\textgreater{} . r\textquotesingle{} . sigma(r).}
\end{Highlighting}
\end{Shaded}

If
\texttt{sigma(rho)\ =\ l\textquotesingle{}\ .\ diamond\ .\ r\textquotesingle{}},
the result is the closed row
\texttt{sigma(l)\ .\ l\textquotesingle{}\ .\ diamond\ .\ r\textquotesingle{}\ .\ sigma(r)}.

\textbf{Lemma 3.3 (Composition).} With this splice definition,
\texttt{(sigma\ o\ tau)(X)\ =\ sigma(tau(X))} for all terms \texttt{X}.

\textbf{Proof.} Dimension terms are immediate. For rows, the only
non-structural case is a row variable at the marker. Applying
\texttt{tau} splices \texttt{tau(rho)} into the outer row; applying
\texttt{sigma} then splices any row variable in \texttt{tau(rho)} at the
same marker and maps the flank dimensions. This is exactly the row-term
component of the composite substitution. \texttt{square}

\textbf{Definition 3.4 (Constraint semantics).} Atomic constraints are:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{t1 = t2}
\NormalTok{t1 \textless{}= t2}
\NormalTok{R1 approx R2}
\NormalTok{R1 \textless{}= R2.}
\end{Highlighting}
\end{Shaded}

A ground substitution \texttt{gamma} satisfies dimension equality by
identity, dimension inequality by Definition 1.2, row equality by flat
equivalence \texttt{approx}, and row inequality by Definition 2.3. Write
\texttt{Sol(Phi)} for the set of ground substitutions satisfying every
constraint in \texttt{Phi}.

The row equality choice is the important one: equality constraints are
marker-blind, while inequality constraints use the marked order.

\textbf{Definition 3.5 (Substitution models and order).} A substitution
\texttt{sigma} models \texttt{Phi}, written \texttt{sigma\ models\ Phi},
iff every ground substitution \texttt{gamma} makes
\texttt{gamma\ o\ sigma} a solution. The substitution preorder is
\texttt{sigma1\ \textless{}=\ sigma2} iff there exists \texttt{u} with
\texttt{u\ o\ sigma1\ =\ sigma2}.

\textbf{Example 3.6 (No principal model).} For
\texttt{Phi\ =\ \{\ 3\_b\ \textless{}=\ alpha\ \}}, the identity
substitution is not a model because \texttt{alpha} could be grounded to
a different atom. The substitutions
\texttt{{[}alpha\ -\textgreater{}\ 3\_b{]}} and
\texttt{{[}alpha\ -\textgreater{}\ 1\_empty{]}} are both models, and
neither factors through the other: ground dimensions cannot be rewritten
by a later substitution. Thus solvable constraint sets need not have a
least substitution model. The solver must return a substitution plus a
residual bound store.

\subsubsection{3.1 Constraint
Generation}\label{31-constraint-generation}

The core operation rules generate row inequalities for ordinary
broadcasting and row equalities for spec-based operations. Let
\texttt{C} be the result shape; \texttt{A}, \texttt{B}, and \texttt{D}
operands; and \texttt{S\_k} the row of kind
\texttt{k\ in\ \{Batch,\ Input,\ Output\}}.

\begin{longtable}[]{@{}ll@{}}
\toprule\noalign{}
Operation & Generated row constraints \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
transpose & \texttt{C\_B\ \textless{}=\ A\_B},
\texttt{C\_I\ \textless{}=\ A\_O}, \texttt{C\_O\ \textless{}=\ A\_I} \\
pointwise unary & \texttt{C\_k\ \textless{}=\ A\_k} for every kind
\texttt{k} \\
pointwise binary & \texttt{C\_k\ \textless{}=\ A\_k},
\texttt{C\_k\ \textless{}=\ B\_k} for every kind \texttt{k} \\
pointwise ternary & result row below each operand row for every kind \\
compose & \texttt{A\_I\ \textless{}=\ B\_O},
\texttt{C\_B\ \textless{}=\ A\_B}, \texttt{C\_B\ \textless{}=\ B\_B},
\texttt{C\_I\ \textless{}=\ B\_I}, \texttt{C\_O\ \textless{}=\ A\_O} \\
compose accumulate & compose constraints plus
\texttt{C\_k\ \textless{}=\ D\_k} \\
batch slice & \texttt{(s\ .\ C\_B)\ approx\ A\_B},
\texttt{C\_I\ approx\ A\_I}, \texttt{C\_O\ approx\ A\_O} \\
permute & \texttt{C\_k\ approx\ T\_lhs,k},
\texttt{T\_rhs,k\ approx\ A\_k} \\
einsum & \texttt{C\_k\ approx\ T\_lhs,k},
\texttt{T\_rhs,i,k\ approx\ A\_i,k} \\
\end{longtable}

This table matches the implementation in \texttt{tensor/shape.ml}:
ordinary pointwise and compose rules emit \texttt{Row\_ineq}, while
batch slice, permute, and einsum emit \texttt{Row\_eq}. Batch slice is
the canonical source of a non-empty leading flank; it prepends the slice
axis to the result batch row.

\subsection{4. The Solver}\label{4-the-solver}

\textbf{Definition 4.1 (Configurations).} A configuration is

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{\textless{} Phi ; sigma ; B \textgreater{}}
\end{Highlighting}
\end{Shaded}

where \texttt{Phi} is the pending constraint list, \texttt{sigma} is the
idempotent substitution, and \texttt{B} is a bound store. The store
records, for dimension variables, atom lower bounds and variable
adjacencies; for row variables, row caps and row adjacencies. A failure
configuration is written \texttt{fail}.

Binding a variable moves the equality into \texttt{sigma}, substitutes
through \texttt{Phi} and \texttt{B}, and re-emits all stored bounds for
the bound variable, instantiated at its new value. This single
re-emission discipline is the reason stored facts can be interpreted as
constraints in the metatheory.

\subsubsection{4.1 Dimension Rules}\label{41-dimension-rules}

\begin{itemize}
\tightlist
\item
  \textbf{DE-refl:} \texttt{t\ =\ t}; discard.
\item
  \textbf{DE-clash:} distinct ground dimensions; \texttt{fail}.
\item
  \textbf{DE-bind:} \texttt{alpha\ =\ t}, \texttt{t\ !=\ alpha}; bind
  \texttt{alpha\ -\textgreater{}\ t}.
\item
  \textbf{DI-top:} \texttt{t\ \textless{}=\ 1\_empty}; discard.
\item
  \textbf{DI-refl:} \texttt{t\ \textless{}=\ t}; discard.
\item
  \textbf{DI-ground:} ground
  \texttt{d\ \textless{}=\ d\textquotesingle{}}; check Definition 1.2,
  else \texttt{fail}.
\item
  \textbf{DI-pin:} \texttt{alpha\ \textless{}=\ d}, with \texttt{d} an
  atom; replace by \texttt{alpha\ =\ d}.
\item
  \textbf{DI-pin-top:} \texttt{1\_empty\ \textless{}=\ alpha}; replace
  by \texttt{alpha\ =\ 1\_empty}.
\item
  \textbf{DI-cap:} \texttt{d\ \textless{}=\ alpha}, with \texttt{d} an
  atom; record the lower bound. If a distinct atom is already recorded,
  bind \texttt{alpha\ -\textgreater{}\ 1\_empty}.
\item
  \textbf{DI-adj:} \texttt{alpha\ \textless{}=\ beta},
  \texttt{alpha\ !=\ beta}; record the adjacency and forward existing
  caps.
\end{itemize}

\textbf{Lemma 4.2 (Dimension rules preserve solutions).} Each
non-failing dimension rule preserves the denoted solution set. Each
failing dimension rule fires only on an unsatisfiable denotation.

\textbf{Proof.} The reflexive and top rules remove valid constraints.
DE-clash and failed DI-ground are unsatisfiable by Definition 1.2.
DE-bind is the standard substitution-preserving move from a pending
equality into \texttt{sigma}.

For DI-pin, the only dimension below an atom is the atom itself. For
DI-pin-top, the only dimension above the top is the top. For DI-cap,
storing a first lower bound only moves the constraint into \texttt{B}.
If a distinct atom is already stored, any common upper bound of the two
atoms is \texttt{1\_empty}, so the two caps are equivalent to
\texttt{alpha\ =\ 1\_empty}. DI-adj only records and forwards the
transitive consequence
\texttt{d\ \textless{}=\ alpha\ \textless{}=\ beta}. \texttt{square}

\subsubsection{4.2 Row Rules}\label{42-row-rules}

Definition 4.3 uses the mnemonic layout from the workshop article. In
all row rules, leading flanks are aligned from the outer-left edge and
trailing flanks from the outer-right edge. Overlapping aligned
dimensions emit the corresponding dimension equalities or inequalities.

\textbf{Definition 4.3 (Row equality rules).}

\begin{itemize}
\tightlist
\item
  \textbf{RE-closed:} both sides closed. Fail iff ranks differ;
  otherwise zip flat rows and emit dimension equalities.
\item
  \textbf{RE-open-closed:} one side open with \texttt{rho}, the other
  closed. Fail iff the open side\textquotesingle s total explicit flank
  length exceeds the closed rank; otherwise bind \texttt{rho} to the
  uncovered middle of the closed flat row, using the inherit-the-split
  placement policy.
\item
  \textbf{RE-nested:} both sides open, distinct variables, and both
  surpluses are on \texttt{rho2}\textquotesingle s side. Bind
  \texttt{rho1\ -\textgreater{}\ m\_l\ .\ \textless{}rho2\textgreater{}\ .\ m\_r}.
\item
  \textbf{RE-cross:} both sides open, distinct variables, with split
  surpluses. Defer the residual word equation
  \texttt{s\ .\ x1\ =\ x2\ .\ t} into closing.
\item
  \textbf{RE-same-empty:} same variable and no surplus; discard.
\item
  \textbf{RE-same-occurs:} same variable and unequal total flank
  lengths; fail.
\item
  \textbf{RE-same-rot:} same variable, equal total flank length, but
  shifted split. Defer the rotational word equation
  \texttt{x\ .\ t\ =\ s\ .\ x} into closing.
\end{itemize}

\textbf{Explanation.} RE-closed is flat equality, so marker placement is
ignored. RE-open-closed forces the flat content of the row variable, but
not its marker placement. The implementation chooses the closed
side\textquotesingle s split when available and the left edge otherwise.
That choice is a policy commitment: sound, but not complete for all
\texttt{approx}-solutions. RE-nested is the same phenomenon with another
row variable in the middle.

RE-cross is a two-variable word equation. Its solutions are the
principal family

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{x2 = s . w}
\NormalTok{x1 = w . t}
\end{Highlighting}
\end{Shaded}

plus finitely many sporadic crossover solutions where \texttt{x2} is a
proper prefix of \texttt{s}. Eagerly binding a fresh middle variable
would keep only the principal family and would be a rank commitment not
entailed by \texttt{approx}. Therefore the correct core rule defers.

RE-same-rot is governed by the classical conjugacy equation
\texttt{x\ .\ t\ =\ s\ .\ x}. Non-empty periodic solutions exist when
\texttt{s} and \texttt{t} are conjugate. The deterministic solver does
not represent that family; it defers until either other constraints
solve the variable or closing chooses the least-material branch.

\textbf{Definition 4.4 (Row inequality rules).}

\begin{itemize}
\tightlist
\item
  \textbf{RI-cap:} operand open and result supplies at least the operand
  flanks. Record the result\textquotesingle s interior residue as a cap
  on the operand row variable.
\item
  \textbf{RI-closed-op:} operand closed and result rank is at least the
  operand flanks. Compare explicit operand material against the result
  flat row; inserted operand middle positions are unconstrained.
\item
  \textbf{RI-short-closed:} result closed and shorter than the
  operand\textquotesingle s known flanks; fail.
\item
  \textbf{RI-deficit:} result open and shorter than operand known flanks
  by \texttt{k\ \textgreater{}\ 0}. Grow the result variable by a
  template with fresh dimensions and a fresh row variable, then
  reprocess.
\end{itemize}

RI-closed-op is easy to state incorrectly. The operand\textquotesingle s
inserted broadcast positions are unconstrained because they are
\texttt{1\_empty}, but the operand\textquotesingle s explicit axes still
pin corresponding result positions, even across structural flanks. The
current implementation performs this flat comparison.

\textbf{Lemma 4.5 (Row rule preservation).} Every semantic row rule
preserves solutions exactly up to extension on fresh variables. Semantic
failures occur only when the denotation has no solution. The placement
and deferred-equation steps are policy steps: every new solution
restricts to an old solution, but some old solutions can be lost.

\textbf{Proof.}

For RE-closed, flat equality of equal-length closed words is equivalent
to the pointwise dimension equalities. Unequal ranks are unsatisfiable.

For RE-open-closed, write the open side as
\texttt{l\ .\ \textless{}rho\textgreater{}\ .\ r} and the closed flat
word as \texttt{w}. The flat equation is
\texttt{gamma(l)\ .\ x\ .\ gamma(r)\ =\ w}, where
\texttt{x\ =\ flat(gamma\ rho)}. Since the flank lengths are fixed, a
solution exists iff
\texttt{\textbar{}l\textbar{}\ +\ \textbar{}r\textbar{}\ \textless{}=\ \textbar{}w\textbar{}};
then \texttt{x} is the unique middle segment and the outer segments emit
exact dimension equalities. Only the split of \texttt{x} around a marker
is policy.

For RE-nested, cancellation of common outer flanks leaves
\texttt{x1\ =\ m\_l\ .\ x2\ .\ m\_r}. The flat content of the binding is
entailed. The inherited split is policy, for the same reason as
RE-open-closed.

For RE-cross, cancellation leaves \texttt{s\ .\ x1\ =\ x2\ .\ t} with
non-empty split surpluses. If
\texttt{\textbar{}x2\textbar{}\ \textgreater{}=\ \textbar{}s\textbar{}},
then \texttt{x2\ =\ s\ .\ w} and \texttt{x1\ =\ w\ .\ t}. If
\texttt{\textbar{}x2\textbar{}\ \textless{}\ \textbar{}s\textbar{}},
then \texttt{x2} is a proper prefix of \texttt{s}; the remaining suffix
of \texttt{s} must be a prefix of \texttt{t}, and \texttt{x1} is forced
to the complementary suffix of \texttt{t}. These cases are exhaustive,
so deferral preserves the denotation. Any later closing choice is
policy.

RE-same-empty removes a reflexive equation. RE-same-occurs with unequal
total flank lengths is unsatisfiable by length. RE-same-rot is a
deferred word equation; deferral is exact, while the eventual
least-material resolution is policy.

For row inequalities, the aligned explicit overlaps are exactly the
pointwise requirements of Definition 2.3. RI-cap packages the remaining
interior requirement as a stored cap. RI-closed-op discards only
positions where the operand expansion inserts \texttt{1\_empty}, and
emits comparisons for explicit operand axes. RI-short-closed violates
flank-fit with a fixed-rank result. RI-deficit is exact because any
solution must give the result variable at least the missing \texttt{k}
axes; those axes decompose uniquely around the marker according to the
template. Fresh dimensions and the fresh row variable are assigned from
that decomposition, and conversely every solution after the template
binding restricts to a solution before it. \texttt{square}

\subsubsection{4.3 Rank-Fact Graph}\label{43-rank-fact-graph}

The row rules without an extra check can diverge on rank-unsatisfiable
inputs:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{rho1 \textless{}= [a] . \textless{}rho2\textgreater{}}
\NormalTok{rho2 \textless{}= [b] . \textless{}rho1\textgreater{}}
\end{Highlighting}
\end{Shaded}

The first constraint forces
\texttt{rank(rho1)\ \textgreater{}=\ rank(rho2)\ +\ 1}; the second
forces the reverse strict inequality. Operationally, each RI-deficit
step grows one row variable and lengthens the other
constraint\textquotesingle s operand side.

\textbf{Definition 4.6 (Rank graph).} Maintain a persistent directed
graph \texttt{G} on row variables. An edge

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{rho {-}{-}k{-}{-}\textgreater{} rho\textquotesingle{}}
\end{Highlighting}
\end{Shaded}

with \texttt{k\ \textgreater{}=\ 0} records the entailed fact

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{rank(gamma rho) \textgreater{}= rank(gamma rho\textquotesingle{}) + k.}
\end{Highlighting}
\end{Shaded}

Edges are recorded at three points:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{RG-bind:} solving
  \texttt{rho\ -\textgreater{}\ l\ .\ \textless{}rho\textquotesingle{}\textgreater{}\ .\ r}
  records
  \texttt{rho\ -\/-(\textbar{}l\textbar{}+\textbar{}r\textbar{})-\/-\textgreater{}\ rho\textquotesingle{}};
\item
  \textbf{RG-equal-flanks:} processing
  \texttt{R\_res\ \textless{}=\ R\_op} with both sides open and equal
  known flank lengths records
  \texttt{rho\_res\ -\/-0-\/-\textgreater{}\ rho\_op};
\item
  \textbf{RG-deficit:} RI-deficit with deficit
  \texttt{k\ \textgreater{}\ 0} and open operand records
  \texttt{rho\_res\ -\/-k-\/-\textgreater{}\ rho\_op} before growing the
  result.
\end{enumerate}

Every insertion is guarded. If the new edge closes a directed cycle of
total positive weight, the solver fails.

No edge is recorded in RI-cap when the result already has surplus known
flanks. That situation entails only
\texttt{rank(rho\_res)\ \textgreater{}=\ rank(rho\_op)\ -\ s}, a
non-positive lower-bound difference. Recording it as a zero or positive
edge is unsound and rejects valid programs.

\textbf{Proposition 4.7 (Exactness for recorded facts).} A finite set of
recorded rank facts is satisfiable over natural-number ranks iff
\texttt{G} has no positive cycle. Therefore the guard fails exactly when
the recorded facts are jointly unsatisfiable.

\textbf{Proof.} If there is a positive cycle, summing the inequalities
around it gives \texttt{rank(rho)\ \textgreater{}=\ rank(rho)\ +\ w} for
\texttt{w\ \textgreater{}\ 0}, impossible.

Conversely, assume there is no positive cycle. For each vertex
\texttt{rho}, define \texttt{r(rho)} as the supremum of the total
weights of directed walks starting at \texttt{rho}. Any walk that
revisits a vertex contains a cycle of weight zero, since all weights are
non-negative and positive cycles are absent. Removing such a cycle does
not decrease the walk weight, so the supremum is attained on a simple
path and is finite. For every edge
\texttt{rho\ -\/-k-\/-\textgreater{}\ rho\textquotesingle{}}, prefixing
that edge to an optimal walk from \texttt{rho\textquotesingle{}} shows
\texttt{r(rho)\ \textgreater{}=\ r(rho\textquotesingle{})\ +\ k}. Thus
\texttt{r} satisfies every recorded fact. \texttt{square}

This proposition proves the guard sound. It does not prove that the
graph records enough facts to catch every possible divergence.

\subsection{5. Solver Metatheory}\label{5-solver-metatheory}

Let the denotation of a configuration be

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{Chat = Phi union eqns(sigma) union constr(B).}
\end{Highlighting}
\end{Shaded}

At row sort, \texttt{eqns(sigma)} is read using \texttt{approx}.

\textbf{Lemma 5.1 (Solution preservation with policy coordinate).} For
every solver step:

\begin{itemize}
\tightlist
\item
  semantic steps preserve \texttt{Sol(Chat)} exactly, up to extension to
  fresh variables;
\item
  semantic failures imply \texttt{Sol(Chat)\ =\ empty};
\item
  policy steps refine the solution set: every new solution is an old
  solution, but not every old solution necessarily survives.
\end{itemize}

\textbf{Proof.} Dimension rules are Lemma 4.2 and row rules are Lemma
4.5. Moving constraints between \texttt{Phi}, \texttt{sigma}, and
\texttt{B} preserves the denotation because the moved equality or bound
is retained in \texttt{eqns(sigma)} or \texttt{constr(B)}, and
re-emitted constraints are the stored constraints instantiated through a
binding already present in the denotation.

The only non-semantic components are exactly:

\begin{itemize}
\tightlist
\item
  marker placement in RE-open-closed and RE-nested;
\item
  closing-stage resolution of RE-cross and RE-same-rot.
\end{itemize}

These choose one representative or branch from a family of
\texttt{approx}-solutions. The committed representative satisfies the
original flat equality, so the step is sound. It can lose solutions
because marked row inequalities can distinguish
\texttt{approx}-equivalent marker placements, and because the deferred
word equations can have non-selected rank branches. \texttt{square}

\textbf{Definition 5.2 (Rank abstraction).} The rank abstraction of
\texttt{Chat} is the finite set of rank facts obtained from row
constraints. For an open-open row constraint comparing

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{l . \textless{}rho\textgreater{} . r}
\NormalTok{l\textquotesingle{} . \textless{}rho\textquotesingle{}\textgreater{} . r\textquotesingle{}}
\end{Highlighting}
\end{Shaded}

as result/operand of \texttt{\textless{}=}, the fact is

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{rank(rho) \textgreater{}= rank(rho\textquotesingle{}) + (|l\textquotesingle{}| + |r\textquotesingle{}|) {-} (|l| + |r|).}
\end{Highlighting}
\end{Shaded}

An equality contributes both directions. Open-closed constraints
contribute the corresponding absolute rank bounds. A rank model is a map
from row variables to natural numbers satisfying these facts.

Every concrete solution induces a rank model. The converse fails because
rank abstraction forgets dimensions and bases.

\textbf{Lemma 5.3 (Rank conservativity).} Every non-failing rule maps a
configuration with a rank model to one with a rank model agreeing on the
old variables and extended to fresh variables.

\textbf{Proof.} Dimension rules do not change row ranks. RE-closed emits
only dimension constraints. RE-open-closed and RE-nested replace a row
equality by a binding whose rank equation is the consumed rank equation.
RE-cross and RE-same-rot defer, so the denotation is unchanged.

RI-cap stores the same rank condition that the original inequality
contributed. RI-closed-op and RI-short-closed are ground-sided; the
non-fail case contributes no new open-open rank obligation. RI-deficit
with deficit \texttt{k} is triggered only when the consumed rank fact
entails \texttt{rank(rho\_res)\ \textgreater{}=\ k}; extend the model by
\texttt{rank(rho\_fresh)\ =\ rank(rho\_res)\ -\ k}. Binding re-emission
substitutes a term whose explicit flank length accounts for exactly the
rank removed from the bound variable, so numeric rank facts are
preserved. \texttt{square}

\textbf{Theorem 5.4 (Termination on rank-satisfiable inputs).} If finite
\texttt{Phi0} has a rank model, then every fair run terminates, with or
without the rank cycle check. The check never fires on such an input.

\textbf{Proof.} Fix a rank model of the initial problem and extend it
along the run using Lemma 5.3. Track each row-inequality lineage through
storage, re-emission, substitution, and reprocessing. If the current
result side of a lineage is
\texttt{l\ .\ \textless{}rho\textgreater{}\ .\ r}, assign the potential
\texttt{psi\ =\ rank(rho)} from the extended rank model; if the result
side is closed, assign \texttt{psi\ =\ 0}.

An RI-deficit step on that lineage with deficit
\texttt{k\ \textgreater{}\ 0} binds the result variable to a template
with a fresh row variable whose rank is \texttt{rank(rho)\ -\ k}, so the
lineage potential strictly decreases. Other row bindings can only
preserve or decrease the potential because their rank facts are entailed
by the model. Therefore each lineage fires RI-deficit finitely many
times.

No row rule forks a row-sort lineage: a consumed row constraint produces
at most one row-sort continuation. Thus the number of lineages is
bounded by the initial finite set plus continuations created by finite
closing commitments; within the solving phase considered here, it is
finite. Since deficit steps are finite, fresh row variables are finite.
Each variable binds at most once, so binding events are finite. Between
binding events, processing pending constraints strictly decreases the
lexicographic measure consisting of the number of pending row
constraints and then dimension constraints, because storing a bound
moves it out of \texttt{Phi} and non-storing rules consume their
trigger. Hence the run terminates.

Every rank-graph edge is an entailed rank fact. The fixed rank model
satisfies all entailed facts, so by Proposition 4.7 no positive cycle
can be recorded. \texttt{square}

\textbf{Lemma 5.5 (Detection Lemma).} Every infinite run inserts, at
some finite stage, an edge that closes a positive cycle in the
persistent rank graph.

\textbf{Proof.} We use the rank abstraction of Definition 5.2 and the
operational facts already used in Theorem 5.4: row-sort lineages are
finite in number and do not fork; every variable binds at most once; and
a fair run eventually processes every re-emitted row inequality.

First, an infinite run must perform infinitely many RI-deficit steps
with open operands. If only finitely many binding events occurred, then
after the last one the pending-constraint measure would strictly
decrease to solved form or failure. Infinitely many bindings require
infinitely many RI-deficit steps, because the equality rules mint only
along existing finite lineages and do not fork them. Deficits against
closed operands are bounded: a closed operand has fixed rank, so after
finitely many result-side growth steps the result is large enough or the
constraint fails. Therefore infinitely many deficits have open operands,
and each records an RG-deficit edge.

Now consider the signed rank abstraction restricted to the finite set of
row-inequality lineages that participate in infinitely many such
deficits. If this restricted abstraction had a rank model, the potential
proof of Theorem 5.4 would bound all deficits on those lineages,
contradiction. Hence it has no rank model. Difference-constraint duality
gives a directed cycle of lineages

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{L\_1, ..., L\_n}
\end{Highlighting}
\end{Shaded}

whose signed weights have positive total \texttt{W\ \textgreater{}\ 0}.
Write the current live variable for the vertex of \texttt{L\_i} as
\texttt{x\_i}, and write \texttt{h\_i\ \textgreater{}=\ 0} for the
amount of explicit rank already peeled off by substitutions from the
cycle\textquotesingle s original variable to \texttt{x\_i}. The
persistent rank graph contains a binding path from the original variable
to \texttt{x\_i} of weight \texttt{h\_i}.

If the original signed weight on edge
\texttt{L\_i\ :\ i\ -\textgreater{}\ i+1} is \texttt{c\_i}, then after
the current substitutions the live weight is

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{c\_i\textquotesingle{} = c\_i + h\_\{i+1\} {-} h\_i.}
\end{Highlighting}
\end{Shaded}

This is just substitution algebra: replacing the source by a template
with \texttt{h\_i} explicit axes subtracts \texttt{h\_i} from the edge
weight, while replacing the target by a template with
\texttt{h\_\{i+1\}} explicit axes adds \texttt{h\_\{i+1\}}. The cycle
sum is invariant:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sum\_i c\_i\textquotesingle{} = sum\_i c\_i = W \textgreater{} 0.}
\end{Highlighting}
\end{Shaded}

Consequently at least one live edge on the cycle is non-negative at
every time. If a live edge has weight zero, fair processing records the
RG-equal-flanks edge. If it has weight \texttt{k\ \textgreater{}\ 0},
fair processing records the RG-deficit edge and then binds the source
variable to a template with \texttt{k} explicit axes. In terms of the
live weights, this sets that edge\textquotesingle s weight to zero and
adds the same \texttt{k} to the preceding edge\textquotesingle s weight.
Thus positive surplus is never destroyed; it is transported backwards
around the finite cycle. Negative slack on an RI-cap edge is finite, so
repeated transport into that edge eventually exhausts the slack and
turns the edge into a zero or positive live edge, which is then
recorded. This is the negative-edge discharge case: a hidden fact
\texttt{rank(x)\ \textgreater{}=\ rank(y)\ -\ s} becomes recorded once
operand-side substitution has inserted at least \texttt{s} axes.

Because the cycle is finite and the run is fair, some full circulation
occurs: for each lineage edge on the positive signed cycle, choose a
time after the previous chosen edge\textquotesingle s target variable
has advanced to the source variable used by the next chosen edge, and at
which the edge is processed with non-negative live weight. Such times
exist by the surplus-transport argument above. At the chosen processing
of edge \texttt{i}, the graph records an edge

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{x\_i\^{}t {-}{-}(c\_i + h\_\{i+1\}\^{}t {-} h\_i\^{}t){-}{-}\textgreater{} x\_\{i+1\}\^{}t.}
\end{Highlighting}
\end{Shaded}

Between this target \texttt{x\_\{i+1\}\^{}t} and the source chosen for
the next edge, the persistent binding graph contains a path of weight
\texttt{h\_\{i+1\}\^{}\{next\}\ -\ h\_\{i+1\}\^{}t}, since bindings only
move from older variables to descendants and offsets only increase.
Composing the recorded lineage edges with these binding paths gives a
directed closed walk in \texttt{G}. Its total weight telescopes:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sum\_i (c\_i + h\_\{i+1\}\^{}\{edge i\} {-} h\_i\^{}\{edge i\})}
\NormalTok{  + sum\_i (h\_i\^{}\{edge i\} {-} h\_i\^{}\{edge i{-}1\})}
\NormalTok{= sum\_i c\_i}
\NormalTok{= W \textgreater{} 0.}
\end{Highlighting}
\end{Shaded}

Thus the closed walk contains a positive cycle. The edge that completes
this walk is guarded by Definition 4.6, so the solver fails at that
finite stage. This contradicts the assumption that the run was infinite
without recording a positive cycle. \texttt{square}

Combining Theorem 5.4 with Lemma 5.5 gives total termination of the core
solving phase on finite inputs: rank-satisfiable inputs terminate by the
potential argument, and rank-unsatisfiable divergence is converted into
finite rank-cycle failure by the persistent graph guard. This is a
termination result for solving, not a completeness result for the later
policy choices: placement commitments, deferred word-equation closure,
and leaf closing can still reject satisfiable \texttt{approx}-systems as
described below.

\textbf{Definition 5.6 (Solved form).} A final configuration
\texttt{\textless{}\ empty\ ;\ sigma\_star\ ;\ B\_star\ \textgreater{}}
is solved when \texttt{sigma\_star} is idempotent, every variable in
\texttt{B\_star} is unsolved, each dimension variable has at most one
atom cap, and all stored row caps and adjacencies have no pending rule
that can fire.

\textbf{Proposition 5.7 (Residual store has a greatest solution).}
Define \texttt{gamma\_up} on unsolved variables by

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{gamma\_up(alpha) = 1\_empty}
\NormalTok{gamma\_up(rho)   = [] . diamond . []}
\end{Highlighting}
\end{Shaded}

and extend through \texttt{sigma\_star}. Then \texttt{gamma\_up}
satisfies \texttt{constr(B\_star)\ union\ eqns(sigma\_star)} and is the
pointwise greatest solution of that final configuration.

\textbf{Proof.} Atom caps \texttt{d\ \textless{}=\ alpha} hold because
\texttt{d\ \textless{}=\ 1\_empty}. Dimension adjacencies hold because
both sides are \texttt{1\_empty}. Row caps and row adjacencies hold
because every row is below the empty row by Proposition 2.6. Bindings
hold by extension through \texttt{sigma\_star}.

For greatestness, any solution maps unsolved dimension variables below
\texttt{1\_empty} and unsolved row variables below the empty row. For
solved variables, use structural induction on the solved term. Ground
dimensions and ground flanks compare by equality; variable occurrences
use the unsolved case; an open row term under \texttt{gamma\_up} has an
empty middle, so any material that another solution inserts at the
marker compares against \texttt{1\_empty} in the expanded middle.
\texttt{square}

\textbf{Corollary 5.8 (Decision status).} Successful solving implies
\texttt{Sol(Phi0)} is non-empty. Semantic failure implies
\texttt{Sol(Phi0)} is empty. Policy failure implies only that the
policy-strengthened problem is empty. On satisfiable,
placement-undiscriminated inputs, every fair run terminates and
succeeds. On rank-unsatisfiable inputs, the Detection Lemma converts the
only possible divergence mode into finite rank-cycle failure.

\textbf{Theorem 5.9 (Representation and most generality,
policy-qualified).} Let a fair run on finite \texttt{Phi0} terminate in
solved form
\texttt{\textless{}\ empty\ ;\ sigma\_star\ ;\ B\_star\ \textgreater{}}.
For the fragment of the run consisting only of semantic steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \texttt{gamma\ in\ Sol(Phi0)} iff \texttt{gamma} extends to a ground
  \texttt{hat\_gamma} over fresh variables such that
  \texttt{hat\_gamma\ =\ hat\_gamma\ o\ sigma\_star} and
  \texttt{hat\_gamma} satisfies \texttt{constr(B\_star)};
\item
  every binding in \texttt{sigma\_star} is entailed by \texttt{Phi0};
\item
  for every substitution \texttt{sigma\ models\ Phi0}, there is
  \texttt{u} with \texttt{u\ o\ sigma\_star\ =\ sigma} on the original
  variables.
\end{enumerate}

For full deterministic runs, these statements hold for the
policy-strengthened system. Up to flat row content, \texttt{sigma\_star}
is still most general; marker placements are the
solver\textquotesingle s deterministic policy, not canonical semantics.

\textbf{Proof.} For semantic steps, Lemma 5.1 gives a chain of exact
solution-preserving rewrites from the initial denotation to the final
denotation. This proves the representation statement. A binding moved
into \texttt{sigma\_star} was introduced only when the corresponding
equality or exact row-content equation was entailed by the current
denotation; composing the preservation chain back to \texttt{Phi0} gives
entailment. If \texttt{sigma\ models\ Phi0}, then every grounding of
\texttt{sigma} is represented by the final configuration. Thus applying
\texttt{sigma} to the residual variables witnesses
\texttt{sigma\ =\ sigma\ o\ sigma\_star} on the original variables.

Policy steps are one-way refinements by Lemma 5.1, so the same proof
applies after adding the policy commitments to the initial system.
\texttt{square}

\textbf{Corollary 5.10 (Principal model recovered when no caps remain).}
If the final bound store contains no atom caps and no row caps, then
\texttt{sigma\_star} itself is a least model of \texttt{Phi0} in the
substitution preorder, modulo the same policy qualification for marker
placement.

\textbf{Proof.} With no caps, the residual store contains only
adjacencies, which are valid under universally quantified free variables
at top or are represented by remaining variables. The representation
theorem therefore says \texttt{sigma\_star\ models\ Phi0}, and Theorem
5.9 gives the factoring property for every other model. \texttt{square}

\subsection{6. Closing}\label{6-closing}

Solving intentionally stops with residual caps. Closing turns a solved
configuration into a total ground substitution. This is a policy stage,
not a principal unification theorem.

\textbf{Definition 6.1 (Closing policy).}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Close leaf variables downward. A leaf dimension gets its saturated
  atom cap if any, otherwise \texttt{1\_empty}; a parameter leaf without
  a cap errors. A leaf row gets the join of the ground parts of its row
  caps, holes filled with \texttt{1\_empty}, and no further axes beyond
  cap extent; parameter holes error.
\item
  Re-solve after those bindings, because stored constraints are
  re-emitted.
\item
  Close all remaining interior variables upward to top
  (\texttt{1\_empty} or the empty row) and re-solve once more.
\end{enumerate}

The row join in step 1 is Proposition 2.7. Hole filling and "no further
axes" are policy commitments.

\textbf{Proposition 6.2 (Closing terminates and is sound).} Every
re-solve launched by a closing commitment terminates. If closing
finishes without \texttt{error} or \texttt{fail}, the resulting total
ground substitution satisfies \texttt{Phi0}.

\textbf{Proof.} A closing commitment binds a variable to a ground value.
After such a binding, split the rank abstraction into variable-variable
facts and facts with at least one ground side. Ground substitution
cannot create new variable-variable rank facts; it only removes
variables from constraints or turns constraints into absolute facts.
Therefore the variable-variable fragment keeps the rank model supplied
by Proposition 5.7 and Lemma 5.3.

The termination proof of Theorem 5.4 needs rank potentials only for
lineages whose result side is still open. Those potentials live in the
variable-variable fragment. Absolute facts can cause finite failure, but
they cannot create the open-open feeding cycle responsible for unbounded
growth. Thus each re-solve has finitely many deficits, finitely many
mints, finitely many bindings, and terminates by the same
pending-constraint measure.

Soundness follows from Lemma 5.1 for each re-solve and from the fact
that each committed ground value is then checked by re-emission of its
stored constraints. If no re-solve fails, all committed values satisfy
the final denotation, which composes back to \texttt{Phi0} along the
sound direction of Lemma 5.1. \texttt{square}

\textbf{Example 6.3 (Incomplete but intentional leaf closing).} Let

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{Phi = \{}
\NormalTok{  3\_b \textless{}= alpha,}
\NormalTok{  5\_b \textless{}= beta,}
\NormalTok{  gamma \textless{}= alpha,}
\NormalTok{  gamma \textless{}= beta}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

with \texttt{alpha} and \texttt{beta} leaves and \texttt{gamma}
interior. The store is satisfiable: \texttt{alpha\ =\ beta\ =\ 1\_empty}
and \texttt{gamma\ =\ 3\_b} is one solution. The deterministic leaf
policy commits \texttt{alpha\ =\ 3\_b} and \texttt{beta\ =\ 5\_b}.
Re-solving then forces \texttt{gamma} below two distinct atoms and
fails. This is a policy rejection, not semantic unsatisfiability. It is
useful behavior for OCANNL because it avoids silently broadening a leaf
tensor with a concrete lower bound unless the user asks for that shape.

\textbf{Proposition 6.4 (Upward closing greatestness).} The
uniform-upward solution \texttt{gamma\_up} of Proposition 5.7 is a
solution of the final configuration. It is pointwise greatest among
solutions of the policy-strengthened system in the marked order.
Moreover, for every solution of the rank-policy-strengthened initial
system, \texttt{gamma(x)\ \textless{}=1\ gamma\_up(x)} at row sort and
\texttt{gamma(x)\ \textless{}=\ gamma\_up(x)} at dimension sort.

\textbf{Proof.} Membership and marked greatestness for the final
configuration are Proposition 5.7. The policy-strengthened initial
system maps exactly to the final configuration by Lemma 5.1 and Theorem
5.9.

For the \texttt{\textless{}=1} statement, dimensions are immediate from
topness or entailed ground bindings. For rows, if \texttt{x} is
unsolved, \texttt{gamma\_up(x)} is the empty row, and every rigidified
flat row is below it. If \texttt{x} is solved by a term \texttt{T}, the
flat content of \texttt{T} is entailed for every solution that respects
the rank-policy commitments. Induct on \texttt{T}. For a closed row,
compare equal-rank rigid flat content pointwise, using the dimension
case. For an open row \texttt{l\ .\ \textless{}rho\textgreater{}\ .\ r},
the rigid flat content under \texttt{gamma} is
\texttt{gamma(l)\ .\ flat(gamma\ rho)\ .\ gamma(r)}, while
\texttt{gamma\_up(T)} is
\texttt{gamma\_up(l)\ .\ diamond\ .\ gamma\_up(r)} with the middle
expanded by \texttt{1\_empty}; flank positions compare by induction and
middle positions compare against \texttt{1\_empty}. \texttt{square}

The unqualified marked-order greatestness over \texttt{Sol(Phi0)} is
false. The single constraint

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{\textless{}rho\textgreater{} approx [] . diamond . [3,5]}
\end{Highlighting}
\end{Shaded}

has a solution with \texttt{rho\ =\ {[}3{]}\ .\ diamond\ .\ {[}5{]}},
incomparable with the implementation\textquotesingle s inherited-split
choice.

\subsubsection{6.1 Marker Provenance and Surface
Discharge}\label{61-marker-provenance-and-surface-discharge}

The abstract calculus admits placement-sensitive counterexamples because
row equality is flat while row inequality is marked. The two-constraint
witness is

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{\textless{}rho\textgreater{} approx [] . diamond . [3,5]}
\NormalTok{[3] . diamond . [9,5] \textless{}= \textless{}rho\textgreater{}.}
\end{Highlighting}
\end{Shaded}

The flat equality permits
\texttt{rho\ =\ {[}3{]}\ .\ diamond\ .\ {[}5{]}}, which satisfies the
inequality, but the deterministic inherited-split policy commits
\texttt{rho\ =\ {[}{]}\ .\ diamond\ .\ {[}3,5{]}}, after which the
inequality fails.

It is tempting to try to prove that this cannot happen for surface
OCANNL programs. That claim should be split into two smaller statements.

\textbf{Syntactic marker provenance.} Every marker position appearing in
a surface-generated constraint has a declared syntactic origin:

\begin{itemize}
\tightlist
\item
  external tensor rows and literal shapes are left-marked closed rows
  (\texttt{beg\_dims\ =\ {[}{]}}, all axes in \texttt{dims});
\item
  spec axes before an ellipsis become leading flanks and axes after an
  ellipsis become trailing flanks;
\item
  batch slice prepends the sliced axis as a leading flank;
\item
  substitution preserves provenance by splicing solved row values at the
  marker;
\item
  closing uses explicit policies: inherited split, cap join, hole fill,
  or upward empty-row close.
\end{itemize}

This is an implementation-checkable invariant, and
\texttt{tensor/shape.ml} supports it: batch slice is the only ordinary
operation that manufactures a non-empty leading flank outside spec
parsing; permute and einsum obtain leading flanks from axes written
before an ellipsis.

\textbf{Surface discharge.} The stronger claim would say that
provenance-respecting constraints never observe the
solver\textquotesingle s marker-placement choice. That is not a purely
formal statement about the current core language. It depends on which
surface programs are considered meaningful, and in particular on whether
a user intended an equality to be a strict inference policy or merely a
broadcast-checking condition. A formal proof can only target a specified
surface fragment.

The defensible version is therefore conditional:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{If a surface fragment uses row equalities only as inference{-}strengthening}
\NormalTok{constraints for spec matching, and never later compares the same inferred}
\NormalTok{flat content under a distinct declared marker split, then placement policy}
\NormalTok{failures cannot occur in that fragment.}
\end{Highlighting}
\end{Shaded}

The proof is by provenance induction. Each equality binding inherits a
marker split from one of the declared origins above. A later marked
inequality can distinguish another split only if the same flat content
is reintroduced with a different declared origin. Pointwise/compose
inequalities reuse tensor-row origins; spec equalities reuse spec
ellipsis origins; slice is the only construct that can deliberately
shift an origin by prepending a leading axis. Thus slice and
axes-before-ellipsis are the cases to audit for any concrete fragment.
Without that fragment restriction, the abstract witness above can be
encoded directly at the row-constraint level and is a real
incompleteness.

\subsection{7. Projection Inference for the
Core}\label{7-projection-inference-for-the-core}

After shape solving and closing, each operation re-derives its own local
constraints with fresh projection identifiers. Projection inference
turns row and dimension constraints into co-iteration classes and fixed
indices for code generation.

The proof in this section covers the core projection language. The
current implementation also contains affine, convolution, and
concatenation handling; those are staged extensions discussed in Section
8.

\textbf{Definition 7.1 (Projection language).} Projection atoms are:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{Proj(p)          an axis projection id}
\NormalTok{Sol(idx)         an externally supplied fixed index or symbol}
\end{Highlighting}
\end{Shaded}

Core equations are:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{Eq(q1, q2)}
\NormalTok{Iter(q).}
\end{Highlighting}
\end{Shaded}

Each projection id has a solved positive size.

\textbf{Definition 7.2 (Elaboration from closed core constraints).}

\begin{itemize}
\tightlist
\item
  \texttt{d1\ =\ d2} emits \texttt{Eq(d1,\ d2)}.
\item
  \texttt{d\_res\ \textless{}=\ d\_op} with \texttt{size(d\_op)\ =\ 1}
  emits \texttt{Eq(d\_op,\ Fix\ 0)}: the operand broadcasts and is
  pinned.
\item
  Other \texttt{d\_res\ \textless{}=\ d\_op} emits
  \texttt{Eq(d\_res,\ d\_op)}.
\item
  \texttt{R1\ approx\ R2} aligns flat rows and emits equality equations
  per pair.
\item
  \texttt{R\_res\ \textless{}=\ R\_op} aligns explicit material from the
  outer edges; surplus result axes are iterated; operand axes of size
  one are pinned to zero as above.
\item
  Every terminal axis emits \texttt{Iter(axis)}.
\end{itemize}

The side condition for fixed indices is
\texttt{0\ \textless{}=\ c\ \textless{}\ size(axis)}.

\textbf{Definition 7.3 (Projection solver).} The core solver maintains:

\begin{itemize}
\tightlist
\item
  a union-find partition of projection ids;
\item
  a partial pin map from classes to fixed indices;
\item
  a set of classes required to iterate.
\end{itemize}

\texttt{Eq(Proj\ p,\ Proj\ q)} unions equal-size classes and fails on
size mismatch. \texttt{Eq(Proj\ p,\ Sol\ i)} pins the class and fails on
conflicting pins. \texttt{Eq(Sol\ i,\ Sol\ j)} checks equality.
\texttt{Iter(Proj\ p)} marks the class. At the end, pinned classes use
their pin; unpinned iterated classes of size greater than one get fresh
iterator symbols; all remaining classes use \texttt{Fix\ 0}.

\textbf{Theorem 7.4 (Projection canonicity).} For a finite core
projection equation set, the partition, pin map, iteration set, and
final labels are independent of processing order. The solver fails
exactly when an equality class contains different sizes, receives
conflicting pins, or contains a false equality between two solved
indices.

\textbf{Proof.} The partition is the least equivalence relation
generated by projection-projection equations, restricted by equal-size
checks. Union-find processing order cannot change the least equivalence
closure. Pins are a partial function on equivalence classes; order can
only change which conflicting pin is observed first, not whether a
conflict exists. The iterate set is the union of all classes mentioned
by \texttt{Iter}, transported through the same equivalence closure.
Labeling is a deterministic function of these three objects, up to
bijective renaming of freshly allocated iterator symbols.
\texttt{square}

\textbf{Definition 7.5 (Product space and reductions).} For every fresh
iterator symbol \texttt{s\_j} assigned to a class of size \texttt{n\_j},
the product space contains the factor \texttt{\{0,\ ...,\ n\_j\ -\ 1\}}.
A factor is a reduction axis iff no component of the left-hand-side
index map uses its iterator.

\textbf{Proposition 7.6 (Coverage and reduction characterization).}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The LHS index map is injective iff every product factor appears in
  some LHS component.
\item
  The LHS index map is surjective onto the LHS index domain iff every
  LHS axis of size greater than one is iterated and no two such LHS axes
  use the same iterator.
\item
  Accumulating read-modify-write is required iff a reduction axis
  exists. Pre-initialization is required iff the LHS map is
  non-surjective, or iff an erasing accumulator is used with a
  non-injective LHS map.
\end{enumerate}

\textbf{Proof.} With only \texttt{Iter} and \texttt{Fix}, two product
points differ exactly in at least one product factor. If a factor
appears in an LHS component, changing it changes the address; if a
factor appears nowhere on the LHS, two product points differing only
there map to the same LHS address. This proves injectivity and the
reduction characterization.

For surjectivity, an LHS axis of size greater than one cannot be covered
by \texttt{Fix\ 0}; it must be labeled by an iterator ranging over the
same size. If two LHS axes share an iterator, the image is diagonal and
misses off-diagonal addresses. If every non-unit LHS axis has a distinct
iterator, every LHS address is realized by choosing those iterator
values and arbitrary values for non-LHS factors. The initialization
statement follows from whether every LHS cell is written exactly once,
at least once, or potentially multiple times. \texttt{square}

\textbf{Lemma 7.7 (Local re-solve and locality, core statement).} If
global shape solving and closing succeed, then for each core operation
the per-operation re-derivation against the closed participating shapes
has a solution, and all spec variables are eliminated. The resulting
projection equivalence depends only on that operation\textquotesingle s
re-derived constraints.

\textbf{Proof.} The operation\textquotesingle s local constraints are
the same schema that contributed to the global constraints, but with
closed shapes and fresh projection identifiers. Restrict the successful
global ground substitution to the participating shapes and rename the
local spec variables accordingly; this is a solution of the local shape
constraints. At closed shapes, flat row equalities have fixed rank and
row variables in specs are pinned by their surrounding concrete rows.
Dimension label variables are then unified with closed operand axes.
Locality follows because projection identifiers are freshened per
operation, and no constraint from another operation mentions them.
\texttt{square}

The proof above is complete for the mathematical core. The
implementation bookkeeping in \texttt{tensor/shape.ml} should still be
audited line-by-line for the full staged language; this is one of the
remaining engineering proof obligations, not a known bug.

\textbf{Theorem 7.8 (Projection soundness).} For a closed core operation
whose local projection solve succeeds:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  every generated tensor address is within the solved shape bounds;
\item
  co-iteration classes are size-uniform;
\item
  co-iteration is local to the operation.
\end{enumerate}

\textbf{Proof.} For an iterator class, Theorem 7.4 only creates a
product factor with the class size; all class members have that size by
the failure condition of union. Thus
\texttt{0\ \textless{}=\ s\_j\ \textless{}\ size(axis)} for every
iterated axis using that symbol. Pinned \texttt{Fix\ c} indices satisfy
the side condition in Definition 7.2, and unpinned uniterated classes
use \texttt{Fix\ 0}, which is valid for every positive axis size.

Size-uniformity is exactly the successful projection-projection union
check. Leastness of co-iteration follows because the partition is the
least equivalence generated by this operation\textquotesingle s
equations. Locality is Lemma 7.7: fresh projection ids prevent
constraints from other operations from entering the closure.
\texttt{square}

\subsection{8. Staged Extensions and Remaining Proof
Obligations}\label{8-staged-extensions-and-remaining-proof-obligations}

The core excludes several constructs used by OCANNL\textquotesingle s
implementation. The current status is:

\textbf{Proposition 8.1 (Padding fold canonicity).} The padding margin
accumulated for an axis is the maximum over finitely many
operation-local padding demands. Because \texttt{max} is associative,
commutative, and idempotent, the accumulated margin is independent of
processing order. \texttt{square}

\textbf{Open 8.2 (Affine and convolution projection determinacy).} The
implementation evaluates affine and convolution projection terms after
the core classes are known. The expected proof is stratified: core
classes are canonical by Theorem 7.4, derived affine terms are
deterministic functions of those classes, and deferred affine equality
checks are pure checks. The proof still needs the exact equations for
valid convolution and size-preserving padding modes and an invariant
saying generation, stored terms, and final checking use the same mode.

\textbf{Open 8.3 (Concat projection determinacy).} The implementation
has a connected-component style coupling for \texttt{Concat}
projections. A proof should show that structural concat equations
generate an order-independent closure and that concat targets receive
deterministic composite indices after their components are solved. This
is plausible from the current \texttt{solve\_proj\_equations} structure,
but it is not covered by the core Theorem 7.4.

\textbf{Open 8.4 (Two-unit shape closing).} Full OCANNL has discardable
variables that can close to a size-zero additive unit, while ordinary
broadcast holes close to \texttt{1\_empty}. Soundness should follow if
the discardability test proves that every zero-closed occurrence is
projected away. The interaction with row instantiation remains to be
formalized.

\textbf{Open 8.5 (Relative completeness of staged solving).}
Unrestricted completeness is not expected. The right theorem is
relative: characterize the surface fragment where stage triggers guess
only forced or solution-irrelevant values, and exhibit counterexamples
just outside it.

\subsection{9. Counterexamples to Broad
Completeness}\label{9-counterexamples-to-broad-completeness}

Rather than trying to prove a broad relative-completeness theorem, the
current evidence points to a boundary: OCANNL\textquotesingle s
deterministic policies are useful, but there are small programs or
solver inputs where the intended solution is clear and the solver either
rejects it or requires an additional user constraint.

\subsubsection{9.1 Periodic Row-Permutation
Shape}\label{91-periodic-row-permutation-shape}

The probe in \texttt{test/einsum/test\_inference\_counterexamples.ml}
compares two programs:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{let}\NormalTok{\%op x = \{ x = uniform1 (); o = [ }\DecValTok{3}\NormalTok{; }\DecValTok{5}\NormalTok{; }\DecValTok{3}\NormalTok{ ] \} }\KeywordTok{in}
\KeywordTok{let}\NormalTok{\%op y = x + (x ++ }\StringTok{"a,b,..r.. =\textgreater{} ..r..,b,a"}\NormalTok{ [ }\StringTok{"a"}\NormalTok{; }\StringTok{"b"}\NormalTok{; }\StringTok{"r"}\NormalTok{ ])}
\end{Highlighting}
\end{Shaded}

and

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{let}\NormalTok{\%op x = \{ x = uniform1 () \} }\KeywordTok{in}
\KeywordTok{let}\NormalTok{\%op y = x + (x ++ }\StringTok{"a,b,..r.. =\textgreater{} ..r..,b,a"}\NormalTok{ [ }\StringTok{"a"}\NormalTok{; }\StringTok{"b"}\NormalTok{; }\StringTok{"r"}\NormalTok{ ]) }\KeywordTok{in}
\NormalTok{Shape.set\_dim a }\DecValTok{3}\NormalTok{;}
\NormalTok{Shape.set\_dim b }\DecValTok{5}
\end{Highlighting}
\end{Shaded}

The concrete program succeeds: the input has shape \texttt{{[}3,5,3{]}},
so the row variable can be \texttt{..r..\ =\ {[}3{]}} and the
permutation is shape-preserving. The abstract-parameter version
currently fails with
\texttt{Shape\_error:\ solved\ dimensions\ for\ axis:\ mismatch}, even
though the same solution is evident: infer the parameter shape as
\texttt{{[}3,5,3{]}}.

This is not a marker-provenance failure. It is the RE-same-rot policy
showing through a surface program. The equation induced by adding the
tensor to its cyclic permutation has a non-empty periodic solution for
\texttt{..r..}; the deterministic solver\textquotesingle s
least-material closure does not search that periodic family.

A user can make the program inferable by providing more shape
information, for example by using a concrete input shape or otherwise
pinning the row. That is exactly the intended relative-completeness
boundary: the code conveys a natural intent, but the current inference
policy does not solve the associated word equation.

\subsubsection{9.2 Leaf Downward
Closing}\label{92-leaf-downward-closing}

The lower-level probe in \texttt{test/einsum/test\_closing\_order.ml}
contains the dimension store

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{3\_b \textless{}= alpha}
\NormalTok{5\_b \textless{}= beta}
\NormalTok{gamma \textless{}= alpha}
\NormalTok{gamma \textless{}= beta}
\end{Highlighting}
\end{Shaded}

with \texttt{alpha} and \texttt{beta} treated as leaves. The store is
satisfiable by raising at least one leaf to the broadcast top, for
example \texttt{alpha\ =\ beta\ =\ 1\_empty}. Deterministic leaf closing
instead commits the capped leaves downward to \texttt{alpha\ =\ 3\_b}
and \texttt{beta\ =\ 5\_b}, then fails when \texttt{gamma} is forced
below two incompatible atoms.

This is not a likely tensor-literal program when the leaf shapes come
from actual arrays, because actual array extents are exact. It is
nevertheless a real policy counterexample for generated or
parameter-like leaves whose caps are lower bounds rather than exact
facts. The user\textquotesingle s intent can be clear in contexts such
as "choose a common broadcastable shape for these two partially
specified leaves"; deterministic downward closing chooses parsimony
instead.

\subsubsection{9.3 Cross-Surplus and Placement
Policy}\label{93-cross-surplus-and-placement-policy}

\texttt{test/einsum/test\_row\_self\_reference.ml} records a row-level
mirror cross-surplus case that is flat-satisfiable but rejected by the
implementation because the inherited split and the conservative trailing
guard select the wrong representative. This remains a row-constraint
counterexample rather than a confirmed ordinary \texttt{\%op}
counterexample. To lift it to the surface, one would need a spec or
slice program that both:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  pins a row variable\textquotesingle s flat content through a
  split-surplus equality; and
\item
  later compares the same flat content under a different declared marker
  split.
\end{enumerate}

That is precisely the marker-provenance audit target from Section 6.1.

\subsection{10. Implementation
Correspondence}\label{10-implementation-correspondence}

The current implementation aligns with the core formalism in the
following places:

\begin{itemize}
\tightlist
\item
  \texttt{tensor/row.ml} represents rows as \texttt{beg\_dims},
  \texttt{dims}, and \texttt{bcast}, exactly the two flanks plus marker.
\item
  \texttt{unify\_row} compares closed rows by flattened content
  (\texttt{beg\_dims\ @\ dims}), matching \texttt{approx}.
\item
  Open-vs-closed row equality matches explicit open flanks against the
  closed flat list and then commits an inherited split for the bound row
  variable.
\item
  Shifted same-variable equality and inequality are deferred until
  another constraint solves the variable or stage-6 upward closing
  supplies the empty middle.
\item
  \texttt{solve\_row\_ineq} records a persistent rank graph with
  non-negative edges for RG-bind, RG-equal-flanks, and RG-deficit. It
  deliberately records no edge for the RI-cap surplus case.
\item
  Closed-closed row inequality compares the operand\textquotesingle s
  explicit leading and trailing material against the
  result\textquotesingle s flat row from the outer edges, so shifted
  explicit axes are checked rather than silently accepted.
\item
  \texttt{tensor/shape.ml} emits row inequalities for pointwise,
  transpose, and compose-style broadcasting, and emits row equalities
  for batch slice, permute, and einsum templates.
\item
  Batch slice prepends the slice dimension to \texttt{beg\_dims},
  confirming the formal motivation for leading flanks.
\item
  Projection identifiers are freshened for solved dimensions before
  local projection solving, supporting the locality theorem for the
  core.
\end{itemize}

The code also implements constructs outside this
report\textquotesingle s core: affine dimensions, convolution padding,
concatenation, total element constraints, discardable zero dimensions,
and staged closing heuristics. These are not known to contradict the
core formalism, but the full soundness theorem for them remains future
work as described in Section 8.

\subsection{11. Summary of Proved and Open
Results}\label{11-summary-of-proved-and-open-results}

Proved in this report:

\begin{itemize}
\tightlist
\item
  dimension lattice facts and non-distributivity;
\item
  row partial order, top, joins, and expansion monotonicity;
\item
  incompatibility of flat equality with the marked order;
\item
  two-sorted rigid-row order and one-shot checking relation;
\item
  solution preservation for all semantic dimension and row rules;
\item
  soundness and exactness of the rank graph for recorded facts;
\item
  termination for finite inputs under the persistent rank-cycle guard;
\item
  the Detection Lemma for rank-unsatisfiable divergent runs;
\item
  satisfiability and greatestness of the residual solved store;
\item
  representation and most-generality for semantic runs, policy-qualified
  for deterministic runs;
\item
  termination and soundness of closing;
\item
  core projection canonicity, coverage, and address safety.
\end{itemize}

Open or deliberately incomplete:

\begin{itemize}
\tightlist
\item
  the syntactic marker-provenance invariant is plausible and
  implementation checkable, but the broader surface-discharge claim is
  only meaningful after choosing a precise surface fragment;
\item
  unconditional completeness for RE-cross and RE-same-rot, unless the
  solver is extended with finite branching over the relevant
  word-equation choices;
\item
  full implementation-level audit of per-operation re-derivation for the
  staged language;
\item
  soundness and determinacy proofs for affine, convolution,
  concatenation, total-element, and zero-discard extensions;
\item
  broad relative completeness of the staged closing heuristics; Section
  9 records current counterexamples instead.
\end{itemize}

\end{document}
